---
title: "QE4PE Mixed-effects Modeling Experiments"
author: "Gabriele Sarti"
date: "Generation date: `r format(Sys.time(), '%b %d, %Y - %H:%M:%S')`"
output: 
  html_document:
    toc: true
    code_folding: show
    toc_float: 
        collapsed: false
        smooth_scroll: true
    number_sections: true
---

# Introduction

This document contains the modeling experiments for the QE4PE paper.

This is partially based on a previous analysis for the publications:

> Toral, A., Wieling, M., and Way, A. (2018). Post-editing effort of a novel with statistical and neural machine translation. Frontiers in Digital Humanities.

> Sarti, G., Bisazza, A., Guerberof-Arenas, A., Toral, A. (2022) DivEMT: Neural Machine Translation Post-Editing Effort Across Typologically Diverse Languages. Proceedings of EMNLP 2022.

## Variables used

### Fixed predictors

1. Length of the MT sentence (characters better than words according to the 2018 paper)
2. Proportion of highlighted contents
3. Highlight modality
4. Target language
5. Document category

### Random factors
1. Subject ID
2. Item ID (nested). Document / Segment

### Dependent variables
1. Editing time for the segment
2. Proportion of edited characters
3. Editing quality (est. by XCOMET, or separate analysis on the MQM subset)

# Analysis

## Load libraries and install required packages if not installed yet

```{r message=FALSE}
packages <- c(
    "car", "effects", "ggplot2", "jtools", "lme4",
    "lmerTest", "mgcv", "optimx", "tidyr", "glmmTMB",
    "DHARMa", "performance", "dplyr", "MuMIn"
)
if (length(setdiff(packages, rownames(installed.packages()))) > 0) {
    install.packages(
        setdiff(packages, rownames(installed.packages())),
        repos = "http://cran.us.r-project.org"
    )
}

library(car)
library(effects)
library(ggplot2)
library(jtools)
library(lme4)
library(mgcv)
library(optimx)
library(glmmTMB)
library(tidyr)
library(DHARMa)
library(performance)
library(dplyr)
library(MuMIn)
```

## Load and process data
```{r message=FALSE}
main <- read.csv("~/Documents/projects/qe4pe/qe4pe/data/task/main/processed_main.csv")

# Drop main entries with has_issues == TRUE
main <- main[main$has_issue == "False", ]
```

```{r}
bin_variable <- function(column, max_val, by_val) {
    num_list <- seq(0, max_val, by = by_val)
    breaks <- c(c(-Inf), num_list, c(Inf))
    # val_list is a sequence from 0 to len(num_list) with step 1
    val_list <- seq(1, length(breaks) - 1, by = 1)

    # Convert all values between num_list[i] and num_list[i+1] to val_list[i]
    return(as.numeric(cut(column, breaks = breaks, labels = val_list, right = TRUE, include.lowest = TRUE)))
}

preprocess <- function(df) {
    # Factorize categorical variables
    df$highlight_modality <- relevel(as.factor(df$highlight_modality), ref = "no_highlight")
    df$tgt_lang <- relevel(as.factor(df$tgt_lang), ref = "nld")
    df$wmt_category <- relevel(as.factor(df$wmt_category), ref = "biomedical")

    df$translator_main_id <- as.factor(df$translator_main_id)
    df$segment_id <- as.factor(df$segment_id)

    # Add proportion of highlighted contents and edit ratio
    df$minor_highlight_ratio <- df$num_minor_highlighted_chars / df$mt_num_chars
    df$major_highlight_ratio <- df$num_major_highlighted_chars / df$mt_num_chars

    # Discretize dependent variables
    df$edit_ratio <- round((1 - (df$num_chars_unchanged / df$mt_num_chars)) * 100, 0)
    df$segment_edit_time_binned <- bin_variable(df$segment_edit_time_filtered, 600, 5)

    # Scale continuous predictors
    df$highlight_ratio <- scale((df$num_minor_highlighted_chars + df$num_major_highlighted_chars) / df$mt_num_chars)
    df$mt_num_chars <- scale(df$mt_num_chars, center = TRUE, scale = TRUE)
    df$doc_edit_order <- scale(df$doc_edit_order, center = TRUE, scale = TRUE)

    # Calculate XCOMET quality improvement
    df$mt_xcomet_qe <- round(df$mt_xcomet_qe * 100, 0)
    df$pe_xcomet_qe <- round(df$pe_xcomet_qe * 100, 0)
    df$diff_xcomet_qe <- df$mt_xcomet_qe - df$pe_xcomet_qe

    return(df)
}
main <- preprocess(main)
```

```{r}
x <- main$segment_edit_time_filtered[main$segment_edit_time_filtered < 600]
print(paste0(round(1 - (length(x) / length(main$segment_edit_time_filtered)), 2), " segments with edit time above ", 600))
```

```{r}
par(mfrow = c(2, 2))
hist(main$segment_edit_time_filtered)
hist(main$edit_ratio)
hist(main$diff_xcomet_qe)
```

```{r}
# Visualize the binned data for edit times
table(main$segment_edit_time_binned)
par(mfrow = c(1, 1))
ggplot(main, aes(x = segment_edit_time_binned)) +
    geom_bar() +
    xlab("Binned Segment Edit Time") +
    ylab("Count") +
    ggtitle("Distribution of Binned Dependent Variable") +
    theme_minimal()
```

## Modeling editing time

### Test different modeling strategies

```{r}
lmer_full <- lmer(
    segment_edit_time_binned ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        highlight_modality +
        mt_xcomet_qe +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | segment_id),
    data = main,
    REML = TRUE
)

# 2. Poisson Mixed-Effects Model
poisson_model <- glmmTMB(
    segment_edit_time_binned ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        highlight_modality +
        mt_xcomet_qe +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | segment_id),
    family = poisson(),
    data = main
)
```

```{r}
# 3. Negative Binomial Mixed-Effects Model
nb_model <- glmmTMB(
    segment_edit_time_binned ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        highlight_modality +
        mt_xcomet_qe +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | segment_id),
    family = nbinom2(),
    data = main
)

# 4. Zero-Inflated Poisson Model
zip_model <- glmmTMB(
    segment_edit_time_binned ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        highlight_modality +
        mt_xcomet_qe +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | segment_id),
    ziformula = ~translator_main_id,
    family = poisson(),
    data = main
)
```

### Testing whether zero-inflation improves the model

```{r}
# 5. Zero-Inflated Poisson Model
zinb_model <- glmmTMB(
    segment_edit_time_binned ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        highlight_modality +
        mt_xcomet_qe +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | segment_id),
    ziformula = ~translator_main_id,
    family = nbinom2(),
    data = main
)
```

```{r}
# Model Comparison
# 1. AIC and BIC Comparison
model_comparison <- AIC(
    lmer_full,
    poisson_model,
    nb_model,
    zip_model,
    zinb_model
)
print("AIC Comparison:")
print(model_comparison)
```

### Model diagnostics

```{r}
summary(nb_model)
simulationOutput <- simulateResiduals(fittedModel = zinb_model)

# residuals are normally distributed
plot(simulationOutput)
```

```{r}
# no significant issue with the dispersion, residuals' variance is well modeled.
testDispersion(simulationOutput)
# no evident zero inflation
testZeroInflation(simulationOutput)

model_performance(zinb_model)

# AIC       |      AICc |       BIC | R2 (cond.) | R2 (marg.) |   ICC |   RMSE
#----------------------------------------------------------------------------
# 55311.274 | 55311.577 | 55548.303 |      0.934 |      0.410 | 0.888 | 17.539

# AIC       | Sigma | Score_log | Score_spherical
#-----------------------------------------------
# 55311.274 | 1.409 |    -3.889 |           0.008

# High conditional R2 = good model, but the fixed effects alone explain only 41.3% of the variance.
# Random effects play a critical role in the model, also shown by high ICC

# RMSE 17.5 ~ avg. error of 85s in predicting editing times
```

```{r}
# Get the predicted values for the zero-inflation component
predicted_zeros_nb <- predict(nb_model, type = "response")
predicted_zeros_zinb <- predict(zinb_model, type = "response")

# Compare the distribution of the predicted and observed scores
# Make 1x3 grid with fixed y-axis limits
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(predicted_zeros_nb, ylim = c(0, 5000))
hist(predicted_zeros_zinb, ylim = c(0, 5000))
hist(main$segment_edit_time_binned, ylim = c(0, 5000))
```

### Model without highlight modality

```{r}
nb_model_highlight_modality_onlyinter <- glmmTMB(
    segment_edit_time_binned ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        mt_xcomet_qe +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | segment_id),
    family = nbinom2(),
    data = main
)

nb_model_no_highlight_modality <- glmmTMB(
    segment_edit_time_binned ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        mt_xcomet_qe +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | segment_id),
    family = nbinom2(),
    data = main
)
```
```{r}
nb_model_highlight_modality_nointer <- glmmTMB(
    segment_edit_time_binned ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        mt_xcomet_qe +
        highlight_modality +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | segment_id),
    family = nbinom2(),
    data = main
)
```
```{r}
AIC(nb_model, nb_model_highlight_modality_nointer, nb_model_highlight_modality_onlyinter, nb_model_no_highlight_modality)
anova(nb_model, nb_model_highlight_modality_nointer, nb_model_highlight_modality_onlyinter, nb_model_no_highlight_modality)
```

Models with `highlight_modality` as an interaction factor outperform those without it.
Highlight modality contributes meaningfully only as an interaction factor with other predictors (target language and category).

## Modeling edit ratio

```{r}
gauss_model <- glmmTMB(
    edit_ratio ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        highlight_modality +
        mt_xcomet_qe +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | segment_id),
    family = gaussian(),
    data = main
)
```

```{r}
cor(main %>% select(edit_ratio, mt_num_chars, highlight_ratio, mt_xcomet_qe))
```

```{r}
# Best model found with dredge(zigauss_model, rank='AIC')
zigauss_model <- glmmTMB(
    edit_ratio ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        highlight_modality +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | doc_id / segment_id),
    ziformula = ~ translator_main_id +
        mt_num_chars +
        tgt_lang +
        wmt_category +
        (tgt_lang * translator_main_id),
    family = gaussian(),
    data = main,
    na.action = "na.fail"
)
```

```{r}
summary(zigauss_model)
AIC(gauss_model, zigauss_model)

#           df      AIC
# gauss_model   21 60760.32
# zigauss_model 49 54717.76

# Zero-inflation is very significant for the edit ratio, as the AIC difference is substantial.

simulationOutput <- simulateResiduals(fittedModel = zigauss_model)

par(mfrow = c(1, 1))
# residuals are normally distributed
plot(simulationOutput)
```

```{r}
par(mfrow = c(1, 2))
# no significant issue with the dispersion, residuals' variance is well modeled.
testDispersion(simulationOutput)
```
```{r}
# Evident and well-modeled zero inflation
testZeroInflation(simulationOutput)

model_performance(zigauss_model)
# AIC       |      AICc |       BIC | R2 (cond.) | R2 (marg.) |   ICC |   RMSE |  Sigma
# -------------------------------------------------------------------------------------
# 54717.760 | 54718.361 | 55052.389 |      0.979 |      0.511 | 0.956 | 10.882 | 10.155

# RMSE 10.8 ~ avg. error of 11% editing rate
```

```{r}
# Get the predicted values for the zero-inflation component
predicted_zeros_gauss <- predict(gauss_model, type = "response")
predicted_zeros_zigauss <- predict(zigauss_model, type = "response")

# Compare the distribution of the predicted and observed scores
# Make 1x3 grid with fixed y-axis limits
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(predicted_zeros_gauss)
hist(predicted_zeros_zigauss)
hist(main$edit_ratio)

rmse_gauss <- sqrt(mean((predicted_zeros_gauss - main$edit_ratio)^2))
rmse_zigauss <- sqrt(mean((predicted_zeros_zigauss - main$edit_ratio)^2))
print(paste0("RMSE for GAUSS model: ", rmse_gauss))
print(paste0("RMSE for ZIGAUSS model: ", rmse_zigauss))
```

### Test usefulness of highlight ratio in predicting edit ratio

```{r}
zigauss_model_highlight_ratio_onlyinter <- glmmTMB(
    edit_ratio ~
        mt_num_chars +
        tgt_lang +
        wmt_category +
        highlight_modality +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | doc_id / segment_id),
    ziformula = ~ translator_main_id +
        mt_num_chars +
        tgt_lang +
        wmt_category +
        (tgt_lang * translator_main_id),
    family = gaussian(),
    data = main,
    na.action = "na.fail"
)
```

```{r}
zigauss_model_highlight_ratio_nointer <- glmmTMB(
    edit_ratio ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        highlight_modality +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | doc_id / segment_id),
    ziformula = ~ translator_main_id +
        mt_num_chars +
        tgt_lang +
        wmt_category +
        (tgt_lang * translator_main_id),
    family = gaussian(),
    data = main,
    na.action = "na.fail"
)
```

```{r}
zigauss_model_no_highlight_ratio <- glmmTMB(
    edit_ratio ~
        mt_num_chars +
        tgt_lang +
        wmt_category +
        highlight_modality +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | doc_id / segment_id),
    ziformula = ~ translator_main_id +
        mt_num_chars +
        tgt_lang +
        wmt_category +
        (tgt_lang * translator_main_id),
    family = gaussian(),
    data = main,
    na.action = "na.fail"
)
```

```{r}
AIC(zigauss_model, zigauss_model_highlight_ratio_onlyinter, zigauss_model_highlight_ratio_nointer, zigauss_model_no_highlight_ratio)
anova(zigauss_model, zigauss_model_highlight_ratio_onlyinter, zigauss_model_highlight_ratio_nointer, zigauss_model_no_highlight_ratio)
```

Highlight ratio is a significant predictor of the edit ratio only when combined with highlight modalities.

## Modeling XCOMET quality improvement

```{r}
gauss_model <- glmmTMB(
    diff_xcomet_qe ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        highlight_modality +
        mt_xcomet_qe +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | doc_id / segment_id),
    family = gaussian(),
    data = main,
    na.action = "na.fail"
)

zigauss_model <- glmmTMB(
    diff_xcomet_qe ~
        mt_num_chars +
        highlight_ratio +
        tgt_lang +
        wmt_category +
        highlight_modality +
        mt_xcomet_qe +
        (highlight_modality * tgt_lang) +
        (highlight_modality * wmt_category) +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | doc_id / segment_id),
    ziformula = ~ translator_main_id +
        mt_num_chars +
        tgt_lang +
        wmt_category +
        (tgt_lang * translator_main_id),
    family = gaussian(),
    data = main,
    na.action = "na.fail"
)
```

```{r}
summary(zigauss_model)
AIC(gauss_model, zigauss_model)

#           df      AIC
# gauss_model   21 54675.80
# zigauss_model 48 42370.97

# Zero-inflation is very significant for the MT quality, as the AIC difference is substantial.

simulationOutput <- simulateResiduals(fittedModel = zigauss_model)

# residuals are normally distributed
plot(simulationOutput)
```


```{r}
par(mfrow = c(1, 2))
# no significant issue with the dispersion, residuals' variance is well modeled.
testDispersion(simulationOutput)
```
```{r}
# Evident and well-modeled zero inflation
testZeroInflation(simulationOutput)

model_performance(zigauss_model)
# AIC       |      AICc |       BIC | R2 (cond.) | R2 (marg.) |   ICC |   RMSE |  Sigma
# -------------------------------------------------------------------------------------
# 41581.044 | 41581.671 | 41922.645 |      0.905 |      0.531 | 0.797 | 6.911 | 8.253

# RMSE 6.9 ~ avg. error of 7% quality improvement
```

```{r}
# Get the predicted values for the zero-inflation component
predicted_zeros_gauss <- predict(gauss_model, type = "response")
predicted_zeros_zigauss <- predict(zigauss_model, type = "response")

# Compare the distribution of the predicted and observed scores
# Make 1x3 grid with fixed y-axis limits
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
hist(predicted_zeros_gauss)
hist(predicted_zeros_zigauss)
hist(main$diff_xcomet_qe)

rmse_gauss <- sqrt(mean((predicted_zeros_gauss - main$diff_xcomet_qe)^2))
rmse_zigauss <- sqrt(mean((predicted_zeros_zigauss - main$diff_xcomet_qe)^2))
print(paste0("RMSE for GAUSS model: ", rmse_gauss))
print(paste0("RMSE for ZIGAUSS model: ", rmse_zigauss))
```

### Simplify model

```{r}
zigauss_model_simple <- glmmTMB(
    diff_xcomet_qe ~
        highlight_modality +
        highlight_ratio +
        tgt_lang +
        mt_xcomet_qe +
        (highlight_modality * highlight_ratio) +
        (1 | doc_edit_order) +
        (1 | translator_main_id) +
        (1 | doc_id / segment_id),
    ziformula = ~ translator_main_id +
        mt_num_chars +
        tgt_lang +
        wmt_category +
        (tgt_lang * translator_main_id),
    family = gaussian(),
    data = main,
    na.action = "na.fail"
)

summary(zigauss_model_simple)
AIC(zigauss_model, zigauss_model_simple)

#                      df      AIC
# zigauss_model        49 41581.04
# zigauss_model_simple 41 41571.64
```

Text length and category are not significant predictors of the XCOMET quality improvement. The
starting quality of the MT output is a significant predictor of the quality improvement.